Pseudogapped Mott Criticality from Vacancy--Kondo Tiling: \protect\\ Quantum Tiling Reconstruction of Non-Fermi Liquid Phases

### 2025-12-10 submissions/TilingUploadROPP/Tiling_Project/QTR_PRL.tex
started creating referee response

%\title{Long-range entangled Pseudogapped Mott Criticality: tracking Kondo breakdown via Quantum Tilling Reconstruction from cavity-Kondo defects}

\title{Quantum Tiling Reconstruction of long-range entangled Pseudogapped Mott Criticality from Cavity-Kondo defects}

@@

\author{Abhirup Mukherjee}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}
\author{S. R. Hassan}
\affiliation{The Institute of Mathematical Sciences, C.I.T. Campus, Chennai 600 113, India}
\author{Anamitra Mukherjee}
\affiliation{School of Physical Sciences, National Institute of Science, Education and Research, HBNI, Jatni 752050, India}
\author{N. S. Vidhyadhiraja}
\affiliation{Theoretical Sciences Unit, Jawaharlal Nehru Center for Advanced Scientific Research, Jakkur, Bengaluru 560064, India}
\author{A. Taraphder}
\affiliation{Department of Physics, Indian Institute of Technology Kharagpur, Kharagpur 721302, India}
%
\author{Siddhartha Lal}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}

@@

\begin{abstract}
%We propose a new theoretical framework—centered on the Vacancy–Kondo (VK) model and implemented via a Quantum Tiling Reconstruction (QTR) protocol—to systematically investigate the Mott transition and pseudogap phenomena in strongly correlated lattice systems. This framework is built upon a lattice-compatible, defect-based formulation in which the fundamental unit is a local Kondo-defect: a vacancy-centered site hosting a spin-$\frac{1}{2}$ moment coupled to its neighboring conduction electrons via exchange and hybridization processes.The QTR procedure restores lattice translation symmetry by tiling these Kondo-defect units across the lattice, yielding a translationally invariant many-body Hamiltonian—termed the \textit{VK-Reconstructed Hubbard–Heisenberg Model}. This construction enables momentum-resolved access to spectral and transport observables, all derived from real-space correlation structures.Within this framework, the Mott transition at half-filling is reinterpreted as a two-stage breakdown of lattice coherence. In the first stage, Kondo screening is frustrated by local charge fluctuations, leading to decoherence and the formation of antinodal gaps that disrupt the connectivity of the Fermi surface. In the second stage, a pseudogap metal emerges, effectively described by an emergent two-channel Kondo (2CK) fixed point in the tiled lattice. The resulting Fermi arcs host non-Fermi liquid excitations and terminate on a critical Fermi surface.As the system nears the transition, these arcs shrink toward nodal points, signaling the onset of a singular nodal metallic state accompanied by enhanced spin-flip correlations and long-range entanglement in real space. This identifies the Mott transition as a fundamentally nonlocal and collective quantum phenomenon—well beyond the reach of traditional impurity-based approaches—and positions the VK+QTR framework as a powerful paradigm for capturing strong correlations, emergent momentum-space geometry, and unconventional quantum criticality in lattice systems.
We present a theoretical framework based on the Vacancy–Kondo (VK) model and a symmetry-restoring method called Quantum Tiling Reconstruction (QTR) to study the Mott transition and pseudogap behavior in correlated lattice systems. The starting point is a local Kondo-defect unit—a vacancy-centered spin-$\frac{1}{2}$ coupled to neighbouring sites through hybridization and exchange. By tiling such a unit across the lattice, QTR produces a translationally invariant many-body Hamiltonian, referred to as the VK-Reconstructed Hubbard–Heisenberg Model. This construction keeps the underlying lattice structure intact and allows momentum-resolved analysis using a real-space formulation. The QTR method is general: by changing the choice of defect unit, different lattice models with specific local interactions can be built. In the VK example, the Mott transition proceeds in two steps. First, Kondo screening is weakened by charge fluctuations, leading to antinodal gaps and partial loss of Fermi surface connectivity. Second, a pseudogap metal appears, described by an effective two-channel Kondo process, with low-energy excitations along Fermi arcs. As the system approaches the transition, these arcs narrow toward the nodal points, resulting in a nodal metal with enhanced long-range spin-slip correlations and entanglement. This framework offers a structured way to connect local interaction effects with the behavior of electrons in the full lattice.
\end{abstract}

\maketitle
\par\noindent\textit{Introduction.}
The pseudogap (PG) and strange metal phases observed in the cuprate high-$T_c$ superconductors remain among the most profound challenges in the physics of strongly correlated electrons~\cite{keimer2015quantum}. The PG phase, which displays a characteristic nodal–antinodal dichotomy and the emergence of Fermi arcs~\cite{loeserKapitulnik1996,Norman1998,Hashimoto2014}\textbf{(CITE LAUCHLI PRL, RAJA PRL)}, has been robustly observed in simulations of the two-dimensional Hubbard model~\cite{KyungKotliar2006,MacridinAzevedo2006,WuFerrero2018,anirbanmott2,HilleAndergassen2020}. Nevertheless, key questions remain unresolved: How is the pseudogap connected to the proximate Mott insulating and superconducting phases~\cite{FradkinRevModPhys2015,Kitatani2023,Sorella2023}? How does it evolve across coupling regimes~\cite{HuangDevereaux2018,Fedor2022,SimkovicFerrero2024}? And is the nodal–antinodal dichotomy intrinsic to the Hubbard model~\cite{Hashimoto2014,Schafer2021}? Most fundamentally, the relation between zero-temperature criticality and finite-temperature crossovers remains ambiguous~\cite{White1998,Ido2018,ProustTaillefer2019,Ponsioen2019,Shengtao2021,XuZhang2022}.

In this work, we introduce a new conceptual and computational framework to address these questions based on the \textbf{Vacancy–Kondo (VK)} model and a symmetry-restoring procedure we term \textbf{Quantum Tiling Reconstruction (QTR)}. At the heart of our construction is a localized \textit{Kondo-defect unit}—a spin-$\frac{1}{2}$ degree of freedom placed at a vacancy in the conduction lattice, coupled to neighbouring sites via hybridization and spin exchange. This defect unit serves as the fundamental building block for constructing the full lattice model. Unlike traditional impurity-based approaches, our formulation retains lattice geometry at every step, enabling a direct route from local many-body physics to global momentum-resolved observables.\\
To construct a translationally invariant correlated lattice system, we apply the Quantum Tiling Reconstruction (QTR) protocol to the local Kondo-defect unit, symmetrically placing it at every site of the lattice using a many-body translation operator. This procedure restores lattice symmetry in a controlled and non-perturbative manner, producing the \textit{VK-Reconstructed Hubbard–Heisenberg Model} (VK-RHH): a momentum-compatible, strongly correlated lattice Hamiltonian that encodes both spin and charge dynamics arising from real-space defect physics.

Importantly, this is a general scheme: the framework itself is adaptable to other vacancy-defect units, enabling systematic exploration of correlated phases in a diverse class of translationally invariant lattice Hamiltonians. In contrast to dynamical mean-field theory (DMFT) and its cluster extensions~\cite{georges1996,Hettler2000DCA,Kotliar2001Cellular,Maier2005,Sakai2023}—which approximate lattice self-energies via local or cluster embeddings—our approach derives the full lattice theory constructively from the local Kondo-defect unit. This preserves the geometric and correlation structure of the square lattice without resorting to self-consistent approximations or coarse graining.

The VK+QTR framework directly reveals the microscopic mechanism underlying pseudogap formation and its evolution into a Mott insulating phase in the half-filled VK-RHH model. As we increase the effective interaction strength and enhance local charge fluctuations, Kondo screening becomes increasingly frustrated and the system undergoes a two-stage coherence breakdown: (1) antinodal regions of the Fermi surface become gapped through a correlation-induced Lifshitz transition~\cite{sakai2009evolution}; and (2) a non-Fermi liquid pseudogap phase emerges, characterized by nodal Fermi arcs and critical spectral weight.

This pseudogap phase is not a mere thermal or numerical crossover, but rather arises from a bona fide quantum phase structure intrinsic to the Kondo-defect unit. Through QTR, these local features propagate coherently into the lattice, giving rise to long-range spin-flip correlations and entanglement patterns. The resulting criticality is inherently nonlocal—marking a sharp departure from the local quantum critical paradigm~\cite{Si2001}—and demonstrates the power of the VK+QTR framework in capturing the effects of lattice geometry and coherence loss in strongly correlated systems.

\par\noindent\textit{The Vacancy–Kondo (VK) Defect Unit.}
To explore pseudogap and Mott physics within a real-space framework that retains lattice geometry, we introduce a localized building block we call the \textbf{cavity-Kondo defect unit}, a composite object formed when a lattice site is removed and replaced by a localized orbital with internal interactions and dynamic coupling to surrounding conduction electrons. The defect orbital at the vacancy carries a local Hubbard interaction $U$, prohibiting double occupancy. It hybridizes with neighboring conduction electrons, 
as well as couples directly to the spins of nearby conduction electrons through an explicit exchange term. The VK defect thus serves as a natural building block for modeling both correlated charge dynamics of itinerant electrons and the dynamical screening of emergent local moments. The Hamiltonian describing the Kondo-defect unit is:
\begin{equation}
\label{eq:VK_defect}
\mathcal{H}_\text{VK} = H_{\text{defect}} + H_{\text{defect-coupling}} + H_{\text{bath}},
\end{equation}
where
\begin{equation}
H_{\text{defect}} = -\frac{U}{2} \left( \hat{n}_{d\uparrow} - \hat{n}_{d\downarrow} \right)^2
\end{equation}
describes the interaction at the vacancy-hosted spin center. This localized spin $\vec{S}_d$ couples to its four nearest neighbor conduction sites $c_{Z\sigma}$ through spin-exchange and one-particle hybridization:
\begin{equation}
H_{\text{defect-coupling}} = \frac{J}{2} \sum_{Z} \vec{S}_d \cdot c^\dagger_{Z\alpha} \boldsymbol{\tau}_{\alpha\beta} c_{Z\beta}
- V \sum_{Z,\sigma} \left( c^\dagger_{Z\sigma} c_{d\sigma} + \text{h.c.} \right).
\end{equation}
Here, $\vec{S}_d$ is the spin-$\frac{1}{2}$ operator at the vacancy, and $Z$ runs over the defect’s neighbors.

The surrounding conduction bath is described by:
\begin{equation}
H_{\text{bath}} = \sum_{\mathbf{k},\sigma} \epsilon_\mathbf{k} c^\dagger_{\mathbf{k}\sigma} c_{\mathbf{k}\sigma}
- \frac{W}{2} \sum_{Z} \left( n_{Z\uparrow} - n_{Z\downarrow} \right)^2,
\end{equation}
where $\epsilon_{\mathbf{k}} = -2t(\cos k_x + \cos k_y)$ defines the tight-binding dispersion at half-filling, and the $W$-term captures local spin-charge interactions that can frustrate the screening of the Kondo defect. A key feature of this VK unit is its spatially discrete embedding. Although the defect breaks translation symmetry locally, the Fourier-transformed Kondo coupling, 
\[
J_{\mathbf{k},\mathbf{k}'} = \frac{J}{2} \left[ \cos(k_x - k'_x) + \cos(k_y - k'_y) \right],
\]
captures how the defect scatters Bloch states of the surrounding lattice, encoding spatial anisotropy and enabling analysis of coherence loss and pseudogap formation. The $C_4$ symmetry of $J_{\mathbf{k},\mathbf{k}'}$ renders it suitable for momentum-resolved tiling.

\par\noindent\textit{Quantum Tiling Reconstruction (QTR).}
To extend the localized VK unit into a translationally invariant many-body system, we introduce the \textbf{Quantum Tiling Reconstruction (QTR)} scheme. In this approach, the local couplings—such as hybridization and spin exchange with neighboring sites—are repeated across the lattice following the symmetry and geometry of the underlying structure. This process builds a global lattice Hamiltonian that inherits its dynamics and geometry from the quantum behavior of the single defect unit. In this way, QTR reconstructs momentum-space features and coherence properties of the system from its local building blocks, providing a consistent framework for studying pseudogap formation, spectral weight redistribution, and other emergent many-body phenomena.

Let \( \mathcal{H}_\text{d}(\mathbf{r}_d) \) denote the local defect-unit Hamiltonian with a vacancy centered at position \( \mathbf{r}_d \). The full lattice Hamiltonian generated via Quantum Tiling Reconstruction (QTR) is defined as:
\begin{equation}
\label{eq:QTR_consistent}
\mathcal{H}_\text{QTR} = \sum_{\mathbf{r}} T^\dagger(\mathbf{r}) \mathcal{H}_\text{d}(\mathbf{r}_d) T(\mathbf{r}) - N H_\text{bath},
\end{equation}
where \( T(\mathbf{r}) \) is a many-body translation operator acting on both conduction electrons and localized defect degrees of freedom with $\mathcal{H}_\text{d}$, and \( N \) is the number of tiling units (introduced to avoid double-counting of the conduction bath).\\ When the cavity-Kondo Hamiltonian is used as the seed unit, $\mathcal{H}_\text{d}\equiv \mathcal{H}_\text{VK}$ the QTR construction yields a specific translationally invariant lattice model $\mathcal{H}_\text{QTR}$, the \textit{VK-Reconstructed Hubbard–Heisenberg Model} ($\mathcal{H}_\text{VK-RHH}$):
\begin{equation}
\begin{aligned}
\mathcal{H}_\text{VK-RHH} =\,
&- \frac{\tilde{t}}{\sqrt{\mathcal{Z}}} \sum_{\langle \mathbf{r}_i, \mathbf{r}_j \rangle, \sigma} \left( c^\dagger_{\mathbf{r}_i \sigma} c_{\mathbf{r}_j \sigma} + \text{h.c.} \right)
- \tilde{\mu} \sum_{\mathbf{r}, \sigma} n_{\mathbf{r} \sigma} \\
&+ \frac{\tilde{J}}{\mathcal{Z}} \sum_{\langle \mathbf{r}_i, \mathbf{r}_j \rangle} \vec{S}_{\mathbf{r}_i} \cdot \vec{S}_{\mathbf{r}_j}
- \frac{1}{2} \tilde{U} \sum_{\mathbf{r}} \left( n_{\mathbf{r} \uparrow} - n_{\mathbf{r} \downarrow} \right)^2.
\end{aligned}
\end{equation}

The renormalized lattice parameters $(\tilde{t}, \tilde{\mu}, \tilde{J}, \tilde{U})$ emerge directly from the local coupling constants of the Kondo-defect unit—such as the Kondo exchange $J$, hybridization $V$, charge fluctuation scale $W$, and onsite interaction $U$—as well as the tiling geometry. This construction encodes the interplay of spin and charge correlations within a fully symmetry-restored lattice framework.

Eigenstates of the QTR Hamiltonian obey a generalized many-body Bloch theorem~\cite{stoyanova}, and the construction enables momentum-resolved access to quantities like spectral functions and self-energies. Crucially, correlation functions and entanglement diagnostics computed within the Kondo-defect unit can be consistently mapped to the full lattice through the symmetry-preserving nature of QTR.
\begin{figure}
    \centering
    \includegraphics[width=0.31\linewidth, height=2.45cm]{phaseDiagram.pdf}
    \includegraphics[width=0.67\linewidth]{SF-DOS.pdf}
    \caption{Left: Phase diagram of impurity model at strong coupling in $U$ in terms of competing dimensionless Kondo ($J/t$) and bath correlation ($W/t$) couplings. A pseudogap phase (red, PG) is observed between the local Fermi liquid (pink, LFL) and local moment (black, LM) phases. Center: $k-$space resolved spin-spin correlation $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$ in the pseudogap phase of the impurity model. Antinodal regions are observed to decouple from Kondo screening of the impurity. Right: Upon tiling, this leads to a $k-$space resolved antinodal gap in the electronic density of states of the lattice model, corresponding to Luttinger surfaces of zeros.}
    \label{spinCorr}
\end{figure}

\par\noindent\textit{Pseudogap Formation via Kondo-Defect Breakdown.}
To understand how the pseudogap (PG) phase emerges within the Vacancy–Kondo (VK) model, we analyze the breakdown of Kondo screening in the large-$U$ regime of the Kondo-defect Hamiltonian. This analysis is carried out through a momentum-resolved scaling flow of the Kondo coupling \(J^{(j)}_{{\bf k}_1, {\bf k}_2}\), using the unitary renormalization group (URG) technique~\cite{anirbanurg1} (details in~\cite{suppmat}).

The RG flow equation takes the form:
\begin{equation}
\label{KondoRGequation}
\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2},
\end{equation}
where $\varepsilon_j$ denotes the energy of the shell being decoupled at the $j^\text{th}$ RG step, and the sum is over all momentum states ${\bf q}$ in the particle sector (PS) of the shell—i.e., all filled states at $T = 0$ in the absence of fluctuations. The notation ${\bf \bar q} = {\bf q} + {\boldsymbol \pi}$ refers to the particle-hole conjugate of ${\bf q}$. The interaction term $W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}$ is marginal under RG flow and encodes spin-charge interactions in the conduction background. While the full behavior of the flow equation is understood via numerical evaluation (discussed below), its analytic structure already shows that an \textit{attractive bath interaction} ($W < 0$) frustrates Kondo screening, ultimately driving a transition to a Mott insulating phase~\cite{Mukherjee_2023}.

By tuning the ratio $W/J$ from zero toward negative values (see Fig.~\ref{spinCorr}, left panel), we uncover a sequence of emergent phases arising from the competition between Kondo screening ($J$) and charge fluctuation suppression ($W$):
\begin{enumerate}
    \item For $|W|/J < (|W|/J)_{\text{PG}}$, the Kondo-defect unit exhibits a \textbf{local Fermi liquid} phase in which the entire Fermi surface participates in coherent Kondo screening.
    \item For $(|W|/J)_{\text{PG}} < |W|/J < (|W|/J)_c$, we observe a \textbf{local pseudogap} phase where only the nodal regions remain actively screened, and the antinodal regions become incoherent.
    \item For $|W|/J > (|W|/J)_c$, the system enters a \textbf{local moment phase} with complete breakdown of Kondo screening at the defect site.
\end{enumerate}

This progression is clearly reflected in the ${\bf k}$-resolved spin correlations between the defect and bath, defined as $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$. In the PG regime (Fig.~\ref{spinCorr}, center), $\chi_s(d,{\bf k})$ shows suppressed correlations in the antinodal regions, indicating momentum-selective screening failure.

We interpret $(W/J)_{\text{PG}}$ as the entry point into the pseudogap phase, and $(W/J)_c$ as its termination into the unscreened regime. Upon performing the Quantum Tiling Reconstruction (QTR), these local transitions propagate into the global lattice model. In particular, the $T=0$ Mott transition of the 2D Hubbard–Heisenberg model reconstructed from VK units now proceeds via a well-defined PG phase, characterized by partial gapping of the Fermi surface. The resulting momentum-resolved density of states exhibits spectral suppression in the antinodal zones, while nodal quasiparticles remain gapless (Fig.~\ref{spinCorr}, right panel).

We now present the microscopic mechanism for this pseudogap formation, as it emerges through the scaling flow of Kondo screening in the defect-centered VK unit.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\linewidth]{zerosFlow.pdf}
    \caption{Left, Right and Center: Decoupling of $J_{{\bf k}, {\bf k'}}$ (positive, negative and zeros shown in yellow, purple and green respectively), with ${\bf k}$ (grey circle) corresponding to the node, antinode and a point mid-way between them on the top right arm of the Fermi surface respectively. This is seen via the appearance of zeros (green patches) of $J_{{\bf k}, {\bf k'}}$ for ${\bf k'}$ initially on the nodal regions of adjacent arms, and their subsequent progression towards the antinodes. The decoupling 
    ends with the onset of the pseudogap.
    }
    \label{rgProgression}
\end{figure}
\par\noindent\textit{Unraveling of Kondo Screening in the Kondo-Defect Unit.}
As shown in Fig.~\ref{rgProgression}, the anisotropic breakdown of Kondo screening within the Kondo-defect unit can be tracked by analyzing the vanishing of the spin-flip scattering amplitude $J_{{\bf k}_N, {\bf k}}$, where ${\bf k}_N = (\pi/2, \pi/2)$ is a nodal point and ${\bf k}$ is a general wavevector on the Fermi surface.

At $W/J = 0$, the scattering amplitude $J_{{\bf k}_N, {\bf k}}$ vanishes for ${\bf k}$ located at antinodal points or nearby nodes. As $W/J$ is decreased towards the pseudogap onset value $(W/J)_{\text{PG}}$, the flow of the renormalized coupling under URG reveals an \textit{unraveling} of Kondo coherence: $J_{{\bf k}_N, {\bf k}}$ becomes RG-irrelevant for ${\bf k}$ near adjacent nodal directions, forming a patch of vanishing amplitude (Fig.~\ref{rgProgression}, left). With further decrease of $W/J$, this patch of zeros expands progressively toward the antinodal region, signaling a systematic decoupling of $J_{{\bf k}_1, {\bf k}_2}$ along connections between different Brillouin zone quadrants (Fig.~\ref{rgProgression}, center and right).

At the critical ratio $W/J = (W/J)_{\text{PG}}$, the antinodal wavevectors become fully disconnected from the rest of the Fermi surface in terms of spin-flip interactions, within numerical precision. This behavior constitutes an interaction-driven \textit{Lifshitz transition} in the effective scattering topology, demarcating the onset of the pseudogap (PG) phase characterized by Fermi arcs~\cite{WuFerrero2018}.

Notably, this phase coincides with the emergence of an effective two-channel Kondo (2CK) structure within the Kondo-defect unit. Each channel corresponds to a pair of symmetry-related Fermi arcs on opposite faces of the Fermi surface. This 2CK character is enforced by the symmetry of the Kondo coupling:
\[
J_{{\bf k},{\bf k'}}= -J_{{\bf k}+{\bf Q},{\bf k'}}=-J_{{\bf k},{\bf k'}+{\bf Q}},
\]
with ${\bf Q} = (\pi, \pi)$. As the PG phase develops further, the disconnected arcs shrink toward the nodal points. Their eventual collapse marks the transition to a fully gapped Mott insulating state, as reconstructed in the lattice model through QTR.

>>>>

\title{Pseudogapped Mott Criticality from Vacancy--Kondo Tiling: \protect\\ Quantum Tiling Reconstruction of Non-Fermi Liquid Phases}

@@

\author{Abhirup Mukherjee}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}
\author{S. R. Hassan}
\affiliation{The Institute of Mathematical Sciences, C.I.T. Campus, Chennai 600 113, India}
\author{Anamitra Mukherjee}
\affiliation{1 School of Physical Sciences, National Institute of Science, Education and Research, HBNI, Jatni 752050, India}
\author{N. S. Vidhyadhiraja}
\affiliation{Theoretical Sciences Unit, Jawaharlal Nehru Center for Advanced Scientific Research, Jakkur, Bengaluru 560064, India}
\author{A. Taraphder}
\affiliation{Department of Physics, Indian Institute of Technology Kharagpur, Kharagpur 721302, India}
%
\author{Siddhartha Lal}
\affiliation{%
 Department of Physical Sciences, Indian Institute of Science Education and Research Kolkata, Nadia - 741246, India
}

@@

\begin{abstract}
We propose a new theoretical framework to study pseudogap formation and the Mott transition in strongly correlated systems, based on the Vacancy--Kondo--Heisenberg (VKH) model and a Quantum Tiling Reconstruction (QTR) scheme. A real-space Kondo-defect unit---comprising a localized spin-1/2 at a lattice vacancy---is tiled across a square lattice to yield a translationally invariant Hubbard--Heisenberg-type model. This VKH-Reconstructed Hubbard--Heisenberg model displays a pseudogap phase characterized by Fermi arc non-Fermi liquid excitations and long-range entanglement. Our framework provides a momentum-resolved, lattice-consistent alternative to impurity-based methods for capturing nonlocal quantum criticality.
\end{abstract}

\maketitle
\textit{Introduction.}

The pseudogap (PG) and strange metal phases in high-$T_c$ cuprates remain central puzzles in correlated electron physics~\cite{keimer2015quantum}. Characterized by a nodal–antinodal dichotomy and Fermi arcs~\cite{Norman1998,Hashimoto2014}, the PG phase has been observed in both experiments and simulations of the 2D Hubbard model~\cite{WuFerrero2018,HilleAndergassen2020}. However, its precise relation to the Mott transition, superconductivity, and zero-temperature criticality remains elusive~\cite{FradkinRevModPhys2015,Sorella2023,XuZhang2022}.

We propose a new real-space approach, combining the \textbf{Vacancy–Kondo–Heisenberg (VKH)} model with a \textbf{Quantum Tiling Reconstruction (QTR)} procedure, to access pseudogap physics from a defect-centered lattice-compatible perspective. At its core lies a \textit{Kondo-defect unit}—a spin-$\frac{1}{2}$ placed at a lattice vacancy and hybridized with surrounding conduction electrons via Kondo exchange and charge coupling. Using QTR, we tile this unit across the lattice while restoring translational symmetry, yielding the \textit{VKH-Reconstructed Hubbard–Heisenberg Model} (VKH-RHH). This enables momentum-resolved access to lattice observables while preserving real-space correlation features.

Our framework reveals the PG as a non-Fermi liquid phase arising from frustrated Kondo screening. A two-stage decoherence occurs: first, antinodal gaps open via a correlation-driven Lifshitz transition; second, a nodal arc metal emerges with two-channel Kondo signatures. As the Mott transition is approached, this critical metal exhibits long-range entanglement and spin-flip correlations—marking a fundamentally nonlocal and collective quantum transition beyond the scope of conventional impurity methods~\cite{Si2001}.

\textit{The VKH Kondo-Defect Unit.}

We define a localized impurity model embedded in a 2D square lattice: a spin-$\frac{1}{2}$ moment resides at a vacancy site and couples via Kondo exchange ($J$) and hybridization ($V$) to its four nearest conduction neighbors. The Hamiltonian is $\mathcal{H}_\text{VKH} = H_{\text{defect}} + H_{\text{defect-coupling}} + H_{\text{bath}}$, where the defect term is $H_{\text{defect}} = -\frac{U}{2} ( \hat{n}_{d\uparrow} - \hat{n}_{d\downarrow} )^2$, and the coupling term reads $H_{\text{defect-coupling}} = \frac{J}{2} \sum_Z \vec{S}_{d} \cdot c^\dagger_{Z\alpha} \boldsymbol{\tau}_{\alpha\beta} c_{Z\beta} - V \sum_{Z,\sigma} ( c^\dagger_{Z\sigma} c_{d\sigma} + \text{h.c.})$.

The conduction bath is described by $H_{\text{bath}} = \sum_{\mathbf{k},\sigma} \epsilon_\mathbf{k} c^\dagger_{\mathbf{k}\sigma} c_{\mathbf{k}\sigma} - \frac{W}{2} \sum_Z (n_{Z\uparrow} - n_{Z\downarrow})^2$, with $\epsilon_{\mathbf{k}} = -2t(\cos k_x + \cos k_y)$. Upon Fourier transforming, the Kondo exchange acquires a $C_4$-symmetric momentum dependence: $J_{\mathbf{k},\mathbf{k}’} = \frac{J}{2} [ \cos(k_x - k_x’) + \cos(k_y - k_y’) ]$.

\textit{Quantum Tiling Reconstruction (QTR).}

To restore translational invariance, we tile the Kondo-defect unit over all sites using a many-body translation operator $T(\mathbf{r})$, producing the QTR Hamiltonian:

$\mathcal{H}_\text{QTR} = \sum_{\mathbf{r}} T^\dagger(\mathbf{r}) \mathcal{H}_\text{VKH} T(\mathbf{r}) - N H_{\text{bath}}.$

This yields a VKH-Reconstructed Hubbard–Heisenberg (VKH-RHH) model of the form:

$\mathcal{H}_\text{VKH-RHH} = -\frac{\tilde{t}}{\sqrt{\mathcal{Z}}} \sum_{\langle \mathbf{r}_i, \mathbf{r}j \rangle, \sigma} ( c^\dagger_{\mathbf{r}i \sigma} c_{\mathbf{r}_j \sigma} + \text{h.c.} )
	-	\tilde{\mu} \sum_{\mathbf{r}} n_{\mathbf{r} \sigma}
-	\frac{\tilde{J}}{\mathcal{Z}} \sum_{\langle \mathbf{r}_i, \mathbf{r}j \rangle} \vec{S}_{\mathbf{r}i} \cdot \vec{S}_{\mathbf{r}_j} -\frac{\tilde{U}}{2} \sum_{\mathbf{r}} ( n_{\mathbf{r} \uparrow} - n_{\mathbf{r} \downarrow} )^2.$

The renormalized parameters arise from the local Kondo-defect unit via: $\tilde{t} = t + 2V$, $\tilde{U} = U + W$, $\tilde{J} = 2J$, and $\tilde{\mu} = 2\mu + \eta$. This construction respects all lattice symmetries and retains local correlation structure.

Eigenstates of the QTR Hamiltonian obey a generalized many-body Bloch theorem~\cite{stoyanova}, and the framework grants direct access to momentum-resolved observables—including spectral functions, self-energies, and entanglement measures—derived from real-space defect dynamics.

\begin{figure}
    \centering
    \includegraphics[width=0.31\linewidth, height=2.45cm]{phaseDiagram.pdf}
    \includegraphics[width=0.67\linewidth]{SF-DOS.pdf}
    \caption{Left: Phase diagram of impurity model at strong coupling in $U$ in terms of competing dimensionless Kondo ($J/t$) and bath correlation ($W/t$) couplings. A pseudogap phase (red, PG) is observed between the local Fermi liquid (pink, LFL) and local moment (black, LM) phases. Center: $k-$space resolved spin-spin correlation $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$ in the pseudogap phase of the impurity model. Antinodal regions are observed to decouple from Kondo screening of the impurity. Right: Upon tiling, this leads to a $k-$space resolved antinodal gap in the electronic density of states of the lattice model, corresponding to Luttinger surfaces of zeros.}
    \label{spinCorr}
\end{figure}

\paragraph*{Pseudogap Formation via Kondo-Defect Breakdown.}

To uncover the origin of the pseudogap (PG) phase in the VKH model, we analyze the momentum-resolved renormalization of the Kondo coupling \( J^{(j)}_{{\bf k}_1, {\bf k}_2} \) using the unitary renormalization group (URG)~\cite{anirbanurg1}. The flow equation reads:
\[
\Delta J^{(j)}_{{\bf k}_1, {\bf k}_2} = -\sum_{{\bf q} \in \text{PS}} \frac{J^{(j)}_{{\bf k}_2,{\bf q}} J^{(j)}_{{\bf q},{\bf k}_1} + 4J^{(j)}_{{\bf q}, {\bf \bar q}} W_{{\bf \bar q}, {\bf k}_2, {\bf k}_1, {\bf q}}}{\omega - \frac{1}{2}|\varepsilon_j| + J^{(j)}_{{\bf q}}/4 + W_{{\bf q}}/2},
\]
where $\varepsilon_j$ is the shell energy, ${\bf q}$ spans filled states at $T = 0$, and ${\bf \bar q} = {\bf q} + {\boldsymbol \pi}$ is its particle-hole partner. The $W$-term is RG-marginal, encoding charge fluctuations. Attractive $W < 0$ frustrates Kondo screening and drives a transition to a Mott insulator~\cite{Mukherjee_2023}.

As $W/J$ is tuned from 0 to negative values (Fig.~\ref{spinCorr}, left), the Kondo-defect unit exhibits three regimes: (i) a local Fermi liquid with full Fermi surface screening for $W/J < (W/J)_{\text{PG}}$; (ii) a PG regime for $(W/J)_{\text{PG}} < W/J < (W/J)_c$ where only nodal regions remain screened; and (iii) a local moment phase beyond $(W/J)_c$ with complete screening breakdown. The spin correlation $\chi_s(d,{\bf k}) = \braket{{\bf S}_d\cdot{\bf S}_{\bf k}}$ reflects this evolution, showing strong suppression in the antinodal zones across the PG phase (Fig.~\ref{spinCorr}, center).

QTR maps this local coherence loss onto the global lattice. In the resulting VKH-RHH model, the $T=0$ Mott transition proceeds through an intermediate PG state with momentum-selective gapping: spectral weight vanishes near the antinodes, while nodal quasiparticles persist (Fig.~\ref{spinCorr}, right). The critical scales $(W/J)_{\text{PG}}$ and $(W/J)_c$ thus delineate a genuine PG phase---not a crossover---emerging from frustrated Kondo dynamics.

\begin{figure}[htpb]
    \centering
    \includegraphics[width=\linewidth]{zerosFlow.pdf}
    \caption{Left, Right and Center: Decoupling of $J_{{\bf k}, {\bf k'}}$ (positive, negative and zeros shown in yellow, purple and green respectively), with ${\bf k}$ (grey circle) corresponding to the node, antinode and a point mid-way between them on the top right arm of the Fermi surface respectively. This is seen via the appearance of zeros (green patches) of $J_{{\bf k}, {\bf k'}}$ for ${\bf k'}$ initially on the nodal regions of adjacent arms, and their subsequent progression towards the antinodes. The decoupling 
    ends with the onset of the pseudogap.
    }
    \label{rgProgression}
\end{figure}

\paragraph*{Unraveling of Kondo Screening.}

The anisotropic breakdown of Kondo screening is traced via the spin-flip amplitude $J_{{\bf k}_N, {\bf k}}$ between the nodal point ${\bf k}_N = (\pi/2, \pi/2)$ and general Fermi momenta ${\bf k}$ (Fig.~\ref{rgProgression}). At $W/J = 0$, $J_{{\bf k}_N, {\bf k}}$ vanishes for ${\bf k}$ near antinodes or adjacent nodes. As $W/J$ approaches $(W/J)_\text{PG}$, RG flow under URG shows this amplitude becomes irrelevant for momenta between adjacent nodes, creating patches of zeros. These regions expand toward the antinodes with decreasing $W/J$, indicating systematic decoupling of $J_{{\bf k}_1, {\bf k}_2}$ between Brillouin zone sectors.

At $(W/J)_\text{PG}$, antinodal states become fully disconnected from the screened nodal region, marking a momentum-selective breakdown of coherence and an interaction-driven \textit{Lifshitz transition}. This signals entry into a pseudogap phase with emergent Fermi arcs~\cite{WuFerrero2018}.

This phase coincides with a two-channel Kondo (2CK) fixed point, where each channel corresponds to symmetry-related arcs. The 2CK structure is symmetry-protected:
\[
J_{{\bf k},{\bf k'}} = -J_{{\bf k}+{\bf Q},{\bf k'}} = -J_{{\bf k},{\bf k'}+{\bf Q}}, \quad {\bf Q} = (\pi,\pi).
\]
With further tuning, these arcs shrink to the nodal points and vanish, completing the transition into the Mott insulator as captured by the VKH-RHH lattice model.

\par\noindent\textit{Momentum-Resolved Spectral Weight Redistribution.}
The dynamical process of spectral weight transfer associated with PG formation is illustrated in Fig.~\ref{chargeCorr}(left). In the PG regime, strong charge fluctuations emerge between nodal and antinodal sectors, which selectively remove low-energy spectral weight from the antinodal regions, pushing it to higher energies and effectively gapping out these regions.
